Models, derivatives, response diversity

### 2024-07-05 models_derivatives_response_diversity.Rmd
tidier, added new fig 2b

---
title: "Models, derivatives, response diversity"
author: "Sophie Moreau"
date: "2024-07-05"
output: html_document
editor_options: 
  chunk_output_type: console
---
#2. Data Analysis
##General Additive Models
Running the predictions for all GAMs with the smaller data frames to reduce running 
time.

@@

# use own function predictions() from function_GAM.R to predict species' models
# across all lakes

# predictions take VERY LONG! and two folders are needed to store the 
# DHARMa and prediction output inside: 
# total_models/gam_check
# total_models/predictions
# with three subfolders: temp, temp_test, depth_temp (total_models/predictions/depth_temp)

# all temperature models, no simplificaiton
# predictions(df_1)
# predictions(df_2)
# predictions(df_3)
# predictions(df_4)
# 
# no_exclusion_predictions_all <- list.files(path = "total_models/predictions/temp_test", pattern = ".rds", full.names = TRUE) |>
#   map_dfr(readRDS)
# 
# # save total predictions as RDS
# saveRDS(no_exclusion_predictions_all, "total_models/df_pred_no_exclusion.rds")

# temperature models with simplified models

@@

no_exclusion_predictions_all <- list.files(path = "total_models/predictions/temp", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

# save total predictions as RDS
saveRDS(no_exclusion_predictions_all, "total_models/df_pred_all.rds")

# sensitivity
# predictions_depth_temp(df_1)
# predictions_depth_temp(df_2)
# predictions_depth_temp(df_3)
# predictions_depth_temp(df_4)
# 
# # issue with cottus sp po
# # statistics of the gamms
# total_models <- bind_rows(df_1, df_2, df_3, df_4)
# 
# stats_gamms(total_models)
# 
# #  compile data frame with all predictions from the saved species' model predictions
# 
# depth_temp_predictions_all <- list.files(path = "total_models/predictions/depth_temp", pattern = ".rds", full.names = TRUE) |>
#   map_dfr(readRDS)
# 
# # save total predictions as RDS
# saveRDS(depth_temp_predictions_all, "total_models/depth_temp_pred_all.rds")

@@

####Fig. 2
Overview of all model predictions

@@

# read in the category table and rename categories
species_across_lakes <-  read_xlsx("species_category_across_lakes.xlsx") |> 
  select(-notes) |> 
    mutate(category = ifelse(category == "native", "non_endemic_native", category)) |>
  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
  mutate(category2 = factor("all"))

@@

# EVTL NEW PLOT 2
# change category names for facet_wrap
category_names <- c(
  `endemic` = "endemic",
  `endemic_and_translocated` = "translocated endemic",
  `non_endemic_native` = "non-endemic native",
  `non_endemic_native_and_translocated` = "translocated non-endemic native",
  `non_native` = "non-native"
)

# colors for the plot
mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
                "endemic_and_translocated" = "#F2AD63", "non_endemic_native_and_translocated" = "#F2AD63")

@@

plot_pred <- predictions_categories_rescaled |>
  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
  geom_line() +
  # how can I get the right color?
   # geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), alpha = 0.3) +
  theme_bw(base_size = 16) +
  ylab("Abundance") +
  xlab("Temperature") +
  scale_color_manual(values = mycolors1, guide = NULL) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "bottom",
        legend.title = element_blank())

plot_category_predictions <- plot_pred + facet_grid(~category, labeller = as_labeller(category_names))

plot_category_predictions

# trying to group by category and color
figure_test_2 <- predictions_categories |>
  ggplot(aes(temp, fit)) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(category)), alpha = 0.3) +
  geom_line(aes(color = factor(category))) +
  theme_bw() +
  facet_wrap(~species + category, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
  scale_color_manual(values = mycolors1, aesthetics = c("color", "fill")) +
  xlab("Temperature") +
  ylab("Abundance")

# tiff(paste("total_models/plots/draft_all_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(figure_test_2)
# 
# # Closing the graphical device
# dev.off()

@@

all_predictions
# save the plot as TIFF
# tiff(paste("total_models/plots/all_predictions_models.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(all_predictions)
# 
# # Closing the graphical device
# dev.off()

##Sensitivity testing
NOT CHANGED 
Testing if depth-driven temperature models can adequately fit abundance data compared to
a temperature and depth model. 
Models were compared in the eleven most abundant species.

@@

```
##Derivatives calculation

We calculated derivatives for each species per lake.

@@

###Temp models
```{r}

@@

# Model 4: before simplification

species_list <- df_abundance_re |> 
  # exclusion of simplified models
  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
                        "Lepomis_gibbosus")) |>
  distinct(Species) |> 
  pull(Species)

    saveRDS(derivatives, paste0("total_models/derivatives_no_simplification/derivatives_", i, "_",  j, ".rds"))

@@

species_list <- df_abundance_re |>
  # simplified models
  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                                "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
                        "Lepomis_gibbosus")) |>
  distinct(Species) |>
  pull(Species)

@@

species_list <- df_abundance_re |> 
  # exclusion of simplified models
  filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
                        "Lepomis_gibbosus")) |>
  distinct(Species) |> 
  pull(Species)

@@

@@

###Fig. S2 
Overview of derivatives in each lake community

###Fig. S3

@@

# above 90% = 2.018455e+01
df_quantiles |> 
  filter(.derivative > 2.018455e+01) |> 
  distinct(species)

@@

# save plot as TIFF
tiff(paste("total_models/plots/plot_deriv_percentiles_2.tiff", sep = ""), units="in", width=19, height=12, res=300)

plot(derivative_percentiles_2)

dev.off()

@@

##Calculating thermal response diversity

>>>>

---
title: "Models, derivatives, response diversity"
author: "Sophie Moreau"
date: "2024-07-05"
output: html_document
editor_options: 
  chunk_output_type: console
---
#General Additive Models
Running the predictions for all GAMs with the smaller data frames to reduce running 
time.

@@

# we have three different functions. One predictions() does not simply any models
# predictions_temp() has multiple models that are simplified (see supplementary 
# figures for the disucssion), predictions_depth_temp() models all species with 
# abundance > 20 with both coviarates temperature and depth.

# For downstream analysis for thermal response diversity models from predictions_temp()
# are used.

# temperature models with simplified models 
# predictions take VERY LONG! and two folders are needed to store the 
# DHARMa and prediction output inside: 
# total_models/gam_check
# needs a folder called total_models/predictions/temp

@@

simplified_predictions_all <- list.files(path = "total_models/predictions/temp", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

# save total predictions as RDS
saveRDS(simplified_predictions_all, "total_models/df_pred_all.rds")

# you can directly use the dataframe df_pred_all for downstream analysis

```
###Simplification and temp depth models
```{r}
# both other functions:

# you can also directly use the dataframe df_pred_no_simplification for decisions about
# which models should be simplified 
# and
# depth_temp_pred_all for the sensitivity testing of the temp + depth models

# testing for which models need to be simplified
# needs a folder: total_models/predictions/temp_test

# all temperature models, no simplificaiton
# predictions(df_1)
# predictions(df_2)
# predictions(df_3)
# predictions(df_4)
# 
# no_exclusion_predictions_all <- list.files(path = "total_models/predictions/temp_test", pattern = ".rds", full.names = TRUE) |>
#   map_dfr(readRDS)
# 
# # save total predictions as RDS
# saveRDS(no_exclusion_predictions_all, "total_models/df_pred_no_simplification.rds")

# testing if modelling temp and depth improves the models
# needs a folder: total_models/predictions/depth_temp

# sensitivity
# predictions_depth_temp(df_1)
# predictions_depth_temp(df_2)
# predictions_depth_temp(df_3)
# predictions_depth_temp(df_4)
# 
# # issue with cottus sp po

# # statistics of the gamms with dept and temp
# total_models <- bind_rows(df_1, df_2, df_3, df_4)
# 
# stats_gamms(total_models)
# 
# #  compile data frame with all predictions from the saved species' model predictions
# 
# depth_temp_predictions_all <- list.files(path = "total_models/predictions/depth_temp", pattern = ".rds", full.names = TRUE) |>
#   map_dfr(readRDS)
# 
# # save total predictions as RDS
# saveRDS(depth_temp_predictions_all, "total_models/depth_temp_pred_all.rds")

@@

####Fig. 2b
New figure 2b with categories across all lakes

@@

# read in the category table and rename categories
species_across_lakes <-  read_xlsx("species_category_across_lakes.xlsx") |> 
  select(-notes) |> 
    mutate(category = ifelse(category == "native", "non_endemic_native", category)) |>
  mutate(category2 = factor("all"))

@@

# change category names for facet_wrap
category_names <- c(
  `endemic` = "endemic",
  `non_endemic_native` = "non-endemic native",
  `non_native` = "non-native"
)

# colors for the plot
mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8")

@@

plot_2b <- predictions_categories_rescaled |>
  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
   geom_line(linewidth = 0.5) +
  theme_bw(base_size = 16) +
  ylab("Abundance") +
  xlab("Temperature") +
  scale_color_manual(values = mycolors1, guide = NULL) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "bottom",
        legend.title = element_blank())

plot_2b_final <- plot_2b + facet_grid(~category, labeller = as_labeller(category_names))

plot_2b_final

# save plot as TIFF
# tiff(paste("total_models/plots/plot_2b.tiff", sep = ""), units="in", width=10, height=4, res=300)
# 
# plot(plot_2b_final)
# 
# dev.off()

####Fig. Supplementary predictions plot
```{r}
# also using predictions_categories from chunk above
# trying to group by category and color

str(categories_supp)

categories_supp <- predictions_categories |>
  arrange(category) |> 
  mutate(species = fct_reorder(species, as.numeric(category))) |> 
  ggplot(aes(temp, fit)) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(category)), alpha = 0.3) +
  geom_line(aes(color = factor(category))) +
  theme_bw() +
  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
  scale_color_manual(values = mycolors1, aesthetics = c("color", "fill")) +
  xlab("Temperature") +
  ylab("Abundance")

categories_supp

# tiff(paste("total_models/plots/supplementary_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(categories_supp)
# 
# # Closing the graphical device
# dev.off()

```
##Sensitivity testing
Testing if depth-driven temperature models can adequately fit abundance data compared to a temperature and depth model.

```{r}
temp_only <- readRDS("total_models/df_pred_all.rds") |> 
  mutate(category = "temp")
temp_depth <- readRDS("total_models/depth_temp_pred_all.rds") |> 
  mutate(category = "temp_depth")

sensitivity

@@

all_predictions
# save the plot as TIFF
# tiff(paste("total_models/plots/all_predictions_models.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(all_predictions)
# 
# # Closing the graphical device
# dev.off()
```

@@

```
#Derivatives calculation

##Data preparation
We calculated derivatives for each species per lake.

@@

##Derivatives
```{r}

@@

@@

species_list <- df_abundance_re |>
  # simplified models
  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                        "Cyprinus_carpio","Phoxinus_csikii", "Salmo_trutta")) |>
  distinct(Species) |>
  pull(Species)

@@

species_list <- df_abundance_re |> 
  # exclusion of simplified models
  filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")) |>
  distinct(Species) |> 
  pull(Species)

@@

#Testing derivatives
##Simplification of models
```{r}
# to get the derivatives before simplifying any model: 
# copy all other derivatives to a new folder called total_models/derivatives_no_simplification and then run this code.

# # Model 4: before simplification
# species_list <- df_abundance_re |>
#   # no simplification
#   filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
#                         "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")) |>
#   distinct(Species) |> 
#   pull(Species)
# 
# species_list <- sort(species_list)
# 
# derivatives <- list()
# gam_output <- list()
# model_prediction <- list()
# 
# 
# for (i in species_list) {
#   
#   data <- df_abundance_re |> 
#     filter(Species == i)
#   
#   gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
#                          +  s(fProtocol, bs = 're'), family = ziP())
#   
#   lake_data <- species_lake |>
#     filter(Species == i)
#   
#   lake_list <- distinct(lake_data, Lake) |>
#     pull()
#   
#   unique_lakes <- distinct(data, Lake) |> 
#     pull()
#   
#   for (j in lake_list){
#     
#     data_lake <- species_lake |>
#       filter(Species == i) |>
#       filter(Lake == j)
#    
#     random_lake <- sample(unique_lakes, 1)
# 
#     newdata <- tibble(mean_last_7days = seq(
#       from = min(data_lake$temp, na.rm = TRUE),
#       to = max(data_lake$temp, na.rm = TRUE), length = 200),
#       fLake = factor(random_lake), fProtocol = factor("VERT"))
#     
#     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
#       mutate(fLake = factor(j)) |>
#       mutate(species = factor(i)) |>
#       rename(temp = mean_last_7days)
#     
#     saveRDS(derivatives, paste0("total_models/derivatives_no_simplification/derivatives_", i, "_",  j, ".rds"))
# 
#   }
# }

@@

##Fig. S3ab

@@

# above the 85th percentile

df_quantiles |> 
  filter(.derivative > 2.018455e+01) |> 
  distinct(species)

# 1 Alburnus_arborella    
# 2 Barbatula_sp_Lineage_I
# 3 Cyprinus_carpio       
# 4 Phoxinus_csikii       
# 5 Salmo_trutta

# NEW: upon inclusion of Lepomis gibbosus
# above 90% = 3.013806e+01
df_quantiles |> 
  filter(.derivative > 3.013806e+01) |> 
  distinct(species)

# 1 Alburnus_arborella    
# 2 Barbatula_sp_Lineage_I
# 3 Phoxinus_csikii       
# 4 Salmo_trutta

@@

# save plot as TIFF
# tiff(paste("total_models/plots/plot_deriv_percentiles_2.tiff", sep = ""), units="in", width=19, height=12, res=300)
# 
# plot(derivative_percentiles_2)
# 
# dev.off()

@@

##Fig. S2: Overview derivatives
Overview of derivatives in each lake community

#Calculating thermal response diversity
